Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-22 manuscript.qmd
fix some cross-references

Pathological involvement is available only for the subset of surgical patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. How this consensus is formed is described in detail in appendix XXX. In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses. If a level was not dissected, the most likely state is determined by PET-CT in most cases.

@@

Using {#eq-bilateral-marginal} along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

@@

## Involvement Data Consensus

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table XXX.

The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side shows the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for EARLYorLATE T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

@@

The model has only one parameter, $p_\text{adv.}$, to describe differences in the patterns of involvement between early and advanced T-category. Higher involvement for advanced T-category is described by evolving the state of involvement overr a larger number of time-steps, while the probability of spread per time step, here the $b_v$ parameters, remains constant. Overall, the model captures the differences between early and advanced T-category very well as seen in @Fig-model-prevalences-overall. However, for some situations the observed difference between early vs advanced T-category in the data is greater or smaller than what our model predicts. This has previously been discussed for the involvement of ipsilateral LNL I [cite 2024 graph extension paper]. In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in appendix XXX.

@@

As discussed in section XXX, we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement. This is appropriate for pathologically confirmed involvement but is an approximation for clinically diagnosed involvement, as the latter is - by definition - incapable of detecting occult disease. In principle, the model can be trained while  differentiating between pathologically confirmed and clinically diagnosed involvement, using a lower sensitivity and specificity for clinical diagnosis. This was not done here for two reasons: 1) this simplification allowed us to compare the observed prevalence of involvement to the model's prediction, and thereby investigate the model's capability to describe data with few interpretable parameters; and 2) the literature values for sensitivity and specificity of the diagnostic modalities are inconsistent with some aspects of the observed data. Future work may aim at developing novel concepts to rigorously differentiating between pathologically confirmed and clinically diagnosed involvement.

@@

# Consensus on most likely involvement {.appendix}

@@

# Prevalence of Midline Extension {.appendix}

>>>>

Pathological involvement is available only for the subset of surgical patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. How this consensus is formed is described in detail in @sec-consensus. In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses. If a level was not dissected, the most likely state is determined by PET-CT in most cases.

@@

Using @eq-bilateral-marginal along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

@@

## Involvement Data Consensus {#sec-methods-consensus}

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.

The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for late T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

@@

The model has only one parameter, $p_\text{adv.}$, to describe differences in the patterns of involvement between early and advanced T-category. Higher involvement for advanced T-category is described by evolving the state of involvement overr a larger number of time-steps, while the probability of spread per time step, here the $b_v$ parameters, remains constant. Overall, the model captures the differences between early and advanced T-category very well as seen in @Fig-model-prevalences-overall. However, for some situations the observed difference between early vs advanced T-category in the data is greater or smaller than what our model predicts. This has previously been discussed for the involvement of ipsilateral LNL I [@ludwig_modelling_2023]. In the context of contralateral spread, the prevalence of midline extension is overestimated for early T-category tumors and underestimated for advanced T-category tumors, which is further discussed in @sec-prevalence-midext.

@@

As discussed in @sec-methods-consensus (and in more detail in @sec-consensus), we assumed for model training that the consensus across all diagnostic modalities represents the true state $X_v$ of lymph node involvement. This is appropriate for pathologically confirmed involvement but is an approximation for clinically diagnosed involvement, as the latter is - by definition - incapable of detecting occult disease. In principle, the model can be trained while  differentiating between pathologically confirmed and clinically diagnosed involvement, using a lower sensitivity and specificity for clinical diagnosis. This was not done here for two reasons: 1) this simplification allowed us to compare the observed prevalence of involvement to the model's prediction, and thereby investigate the model's capability to describe data with few interpretable parameters; and 2) the literature values for sensitivity and specificity of the diagnostic modalities are inconsistent with some aspects of the observed data. Future work may aim at developing novel concepts to rigorously differentiating between pathologically confirmed and clinically diagnosed involvement.

@@

# Consensus on most likely involvement {#sec-consensus .appendix}

@@

# Prevalence of Midline Extension {#sec-prevalence-midext .appendix}
